Machine Learning Methods in Functional Programming and Interactive Theorem Proving

### 2015-10-28 TransferReport/report.tex
More work

\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}. It uses off-the-shelf ATP to verify the conjectures of \textsc{QuickSpec}. \textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes. This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}, useful alongside automated test suites.

In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems to prove, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test. There are likely lessons that each can learn from the other's approach to term generation.

@@

Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures. An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform out later exploration of the literature.

@@

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case. Proofs by induction often require generalisation, since the specific obligations which arise during the proof may be unsuitable for the inductive hypotheses which are available. However, we cannot blindly generalise all obligations we encounter, since \emph{over-generalising} results in obligations which are too strong to prove at all. We must therefore rely on heuristics to guide the generation of generalised conjectures, which we can view as a simple form of interestingness.

A classic example of generalisation occurs when we try to reason about \emph{tail-recursive} definitions \cite{kapur2003automatic}. Consider the following addition functions, defined for the Peano naturals \hs{Z :: Nat} and \hs{S :: Nat -> Nat}:

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
plus :: Nat -> Nat -> Nat
plus n  Z    = n
plus n (S m) = S (plus n m)

plus_tail :: Nat -> Nat -> Nat
plus_tail n  Z    = n
plus_tail n (S m) = plus_tail (S n) m
\end{lstlisting}

Both of these functions implement addition, but the \hs{plus_tail} variant is tail recursive: it does not modify the return value obtained through recursion (in the second clause), and hence we can \emph{replace} the caller's stack frame with the callee's, giving constant space usage. rather than pushing a new frame on top. Compare this to the use of recursion in \hs{plus}, where the value gets passed to the \hs{S} constructor; this manipulation requires the original stack frame to be available after the recursive call returns, and hence we must push a new frame for each iteration. ch recursive call, so that we can pop back to and hence we require the stack frame. In practical terms, the recursive call in \hs{plus_tail} can be implemented in constant space, since the we can discard the stack frames
\fi

@@

TODO: https en.wikipedia.org/wiki/Discovery system

@@

TODO: Expand

@@

\emph{Artificial curiosity} (AC) is the drive to seek out input or data which is in some sense ``interesting'' \cite{schmidhuber2006developmental}. As an unsupervised learning task, AC has no access to labels or meanings associated with its data; the only features it can learn are the structure and relationships between observations and actions. The aim of using an ``interestingness'' measure is to force the AC system away from inputs which are not amenable to learning; either because they are simple enough to be completely characterised already, or because they are so complex as to be unintelligible. This property is succinctly captured by the \emph{Wundt curve} TODO: Cite.

@@

TODO: \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
TODO: \cite{Piantadosi.Tenenbaum.Goodman:2012}
TODO: \cite{Wille:2005}
TODO: \cite{colton1999automatic}
TODO: \cite{colton2000agent}
TODO: \cite{colton2012automated}
TODO: \cite{lenat1977automated}
TODO: \cite{mullerunderstanding}
TODO: \cite{zzzzz57269a963d945d8964c69b117f24e24c}
TODO: \cite{Bundy.Cavallo.Dixon.ea:2015}
TODO: \cite{johansson2009isacosy}
TODO: \cite{spector2008genetic}

@@

TODO: \cite{Gartner2003}
TODO: \cite{Oveisi.Oveisi.Erfanian.ea:2012}
TODO: \cite{bakir2007predicting}
TODO: \cite{conf/ijcai/Plate91}
TODO: \cite{goller1996learning}
TODO: \cite{kwasny1995tail}
TODO: \cite{pollack1990recursive}
TODO: \cite{zanzotto2012distributed}
TODO: \cite{zzzzzba6c2484534f9d1105400e6d693a1d3b}

@@

\bibliographystyle{plain}
\bibliography{../Bibtex}

>>>>

\textsc{HipSpec} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}. It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \textsc{QuickSpec}. \textsc{QuickSpec}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \textsc{QuickCheck} counterexample finder, then conjectures equations relating the members of these classes. This approach works well as a lemma generation system, making \textsc{HipSpec} a capable inductive theorem prover as well as a theory exploration system \cite{claessen2013automating}. \textsc{HipSpec} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \textsc{HipSpec} alongside more traditional QA tools like \textsc{QuickCheck}, \textsc{HUnit} and \textsc{Criterion}.

In fact, there are similarities between the way a TE system like \textsc{HipSpec} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \textsc{QuickCheck} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test. There are likely lessons that each can learn from the other's approach to term generation.

@@

Before exploring abstract definitions of interestingness, we can first consider some scenarios which arise during formal proof where we are forced to generate conjectures. An analysis of these situations, and the subsequent theorems they produce, will contribute towards an empirical justification for what is interesting (at least from a utilitarian point of view) and inform our later exploration of the literature.

@@

\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case. Although counterintuitive, a generalised statement can sometimes be \emph{easier} to prove than the original. This arises often in inductive proofs, since the specific obligations which arise in the proof may be incompatible with the available inductive hypotheses.

However, we cannot blindly generalise \emph{all} obligations we encounter, since \emph{over-generalising} results in obligations which are so strong that they are unprovable, or even false. We must therefore rely on heuristics to guide the generation of generalised conjectures, and hence perform a kind of exploration.

\subsubsection{Tail-recursion}

A classic example of generalisation occurs when we try to reason about \emph{tail-recursive} definitions \cite{kapur2003automatic}. Consider the following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
Inductive Nat : Set := Z : Nat
                     | S : Nat -> Nat.

Fixpoint plus      (n m : Nat) := match n with
                                      | Z    => m
                                      | S n' => S (plus n' m)
                                  end.

Fixpoint plus_tail (n m : Nat) := match n with
                                      | Z    => m
                                      | S n' => plus_tail n' (S m)
                                  end.
\end{lstlisting}

\iffalse
Haskell equivalent:

plus :: Nat -> Nat -> Nat
plus      n  Z    = n
plus      n (S m) = S (plus n m)

plus_tail :: Nat -> Nat -> Nat
plus_tail n  Z    = n
plus_tail n (S m) = plus_tail (S n) m
\fi

Both of these functions implement addition, but the \coq{plus_tail} variant is tail-recursive; i.e. it can be implemented in constant space using a loop. However, if we want to \emph{prove} that the definitions are (pointwise) equal, we run into difficulties:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
(* Solve equalities by beta-normalising both sides *)
Ltac triv := try (simpl; reflexivity).

(* Prove equivalence of plus and plus_tail *)
Theorem equiv : forall n m, plus n m = plus_tail n m.
  induction n; triv. (* Base case is trivial *)

  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
  intro m.

  (* Beta-reduce the right-hand-side (justification is trivial) *)
  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.

  (* Use induction hypothesis to replace plus_tail with plus *)
  rewrite <- (IHn (S m)).
\end{lstlisting}

At this point we must prove \coq{plus (S n) m = plus n (S m)}, which seems reasonable. However, if we try to proceed by another induction on \coq{n}, we get stuck with incompatible induction hypotheses. Specifically, the \emph{conclusion} of our new induction hypothesis is exactly the equation we need:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
\end{lstlisting}

Yet we cannot provide it with the argument it needs, as our original induction hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
IHn : forall x, plus (S n') x = plus_tail (S n') x
\end{lstlisting}

We are forced to abandon the proof, despite such a reasonable-looking intermediate goal. In fact, we \emph{can} prove that goal in isolation:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
Lemma gen n m : plus (S n) m = plus n (S m).
  induction n; triv. (* Base case is trivial *)

  (* Move all S constructors outside *)
  simpl. rewrite <- IHn. simpl.

  (* Trivial *)
  reflexivity.
Defined.
\end{lstlisting}

We can now resume our proof of \coq{equiv}, which is easily discharged:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
  rewrite (gen n m).
  reflexivity.
Defined.
\end{lstlisting}

\subsubsection{Associativity}

@@

@@

% TODO: Expand

@@

\emph{Artificial curiosity} (AC) is the drive to seek out input or data which is in some sense ``interesting'' \cite{schmidhuber2006developmental}. As an unsupervised learning task, AC has no access to labels or meanings associated with its data; the only features it can learn are the structure and relationships between observations and actions. The aim of using an ``interestingness'' measure is to force the AC system away from inputs which are not amenable to learning; either because they are simple enough to be completely characterised already, or because they are so complex as to be unintelligible. This property is succinctly captured by the \emph{Wundt curve} (see figure )

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.75]
    %\node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{/home/chris/DELETEME/screenshot.png}};
    \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
    \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
    \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label

    \draw[black,ultra thick] (3, 4)
         to[out=0,   in=240] (5.05, 6)
         to[out=60,  in=180] (6.5,  8.2)
         to[out=0,   in=105] (8.1,  6)
         to[out=-75, in=120] (8.8,  4)
         to[out=-60, in=150] (11.4, 0.8);

    % This version is closer, but a little jagged
    \iffalse
    \draw[black,ultra thick] (3, 4)
         to[out=0,   in=240] (5.05, 6)
         to[out=60,  in=225] (6,    8)
         to[out=45,  in=180] (6.5,  8.2)
         to[out=0,   in=135] (7.2,  8)
         to[out=-45, in=105] (8.1,  6)
         to[out=-75, in=120] (8.8,  4)
         to[out=-60, in=150] (11.4, 0.8);
    \fi
\end{tikzpicture}

\caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}. The author describes ``hedonic value'' and ``arousal potential'' as terms meant to cover \textquote{both reward value\dots and preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}

\label{wundt}
\end{figure}

% TODO: Cite.

@@

TODO: \cite{Montano-Rivas.McCasland.Dixon.ea:2012}
TODO: \cite{Piantadosi.Tenenbaum.Goodman:2012}
TODO: \cite{Wille:2005}
TODO: \cite{colton1999automatic}
TODO: \cite{colton2000agent}
TODO: \cite{colton2012automated}
TODO: \cite{lenat1977automated}
TODO: \cite{mullerunderstanding}
TODO: \cite{Bundy.Cavallo.Dixon.ea:2015}
TODO: \cite{johansson2009isacosy}
TODO: \cite{spector2008genetic}
TODO: https en.wikipedia.org/wiki/Discovery system

@@

TODO: \cite{Gartner2003}
TODO: \cite{Oveisi.Oveisi.Erfanian.ea:2012}
TODO: \cite{bakir2007predicting}
TODO: \cite{conf/ijcai/Plate91}
TODO: \cite{goller1996learning}
TODO: \cite{kwasny1995tail}
TODO: \cite{pollack1990recursive}
TODO: \cite{zanzotto2012distributed}

@@

\bibliographystyle{plain}
\bibliography{/home/chris/Documents/ArchivedPapers/Bibtex}
